---
title: "Earthquake Risk"
author: "Sonal Agarwal, Wafaa Lawai"
date: "2026-03-09"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Downloading libraries
```{r}
library(terra)
library(readr)
```

Downloading all our variables 

```{r}
#liquefaction zones
liquefaction <- vect("Soil Liquefaction Hazard Zone_20260223/geo_export_dfa19071-7e4a-4377-a4e6-7b26faeb6a64.shp")

#alameda county boundary
alameda <- vect("County_Boundary_-2312687860649160244/County_Boundary.shp")

#Fault lines
faults <- vect("faults/shapefiles/FAM_Qt_Faults.shp")

#population data for Alameda County 
pop <- read.csv("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/alameda_pop.csv", header = FALSE)

#Census tracts for Alameda County
tracts <- vect("Census_Tract_Boundaries_-812809170864177660.geojson")
```


Getting variables into correct form

1) County boundary 

```{r}

alameda <- project(alameda, "EPSG:6417")
```

2) Liquefaction

Project, Crop, and Mask to Alameda County Boundary
```{r}
liquefaction <- project(liquefaction, alameda)
liquefaction <- crop(liquefaction, alameda)
liquefaction <- mask(liquefaction, alameda)

#plotting for visualization
plot(liquefaction, "liq")
plot(alameda, add = TRUE)

```

Converting liquefaction vector into raster 

```{r}
r <- rast(alameda, nrow = 180, ncol = 360)

liquefaction <- terra::rasterize(liquefaction, r, "liq", background = 2)


```

Filling in background with "safe" value
```{r}
cls <- data.frame(id=0:2, zone=c("H", "VH", "S"))

levels(liquefaction) <- cls

liquefaction <- crop(liquefaction, alameda)
liquefaction <- mask(liquefaction, alameda)

```
Reclassify Raster into numeric values
```{r}

reclass_liquefaction <- matrix(c(
  2, 0,
  1, 0.5,
  0, 1
), ncol = 2, byrow = TRUE)

liquefaction_binned <- classify(liquefaction, reclass_liquefaction)
plot(liquefaction_binned) #this is the final layer that will be summed 
```

3) Tracts

Project tracts and find area of each tract for density calculation
```{r}

tracts <- project(tracts, liquefaction)

#finding area 
tract_area <- expanse(tracts, unit = "m")
tracts$area <- tract_area
```

3) Faults 

```{r}
faults <- project(faults, 'epsg:6417')

alameda_faults <- crop(faults, alameda)
alameda_faults <- mask(faults, alameda)
```

Convert faults vector into a distance raster
```{r}
# createa  new raster to serve as distance raster
r <- rast(alameda, nrow = 180, ncol = 360)

dists <- distance(r, alameda_faults)

plot(dists)
plot(alameda_faults, add = TRUE, col = 'white')
```

Crop distance raster to county boundaries
```{r}
dists <- crop(dists, alameda)
dists <- mask(dists, alameda)
```

Distance raster binning/ reclass
```{r}
distmax = 14002.541
distmin  = 0.0047

dists_range <- (distmax - dists) / (distmax - distmin)
plot(dists_range) #this is the final layer that will be summed
```

4) Population density

Tidy population data
```{r}
#head(pop)

pop <- t(pop)
pop <- as.data.frame(pop)

colnames(pop) <- c("Tract", "Population")
pop <- pop[-1,]
```

Create a separate ID column in the population data (this is necessary for when we merge later)
```{r}
pop$ID <- pop$Tract

for (i in 1:379) {
  pop$ID[i] <- parse_number(pop$ID[i])
  pop$Population[i] <- parse_number(pop$Population[i])
}

pop$Population <- as.numeric(pop$Population)
```

Cleaning area data
```{r}
for (i in 1:378) {
  tracts$DIST_NAME[i] <- parse_number(tracts$DIST_NAME[i])
}
```

Merge population data into spatvector with land area
```{r}
colnames(pop) <- c("Tract", "Population", "DIST_NAME")

merged <- merge(tracts, pop, by = "DIST_NAME")

```

Calculate density

```{r}
merged$density <- merged$Population / merged$area

# visualizing our data
plot(merged, "density", breakby = 'cases')
```

Turn this into a raster

```{r}
tracter <- rasterize(merged, r, "density")
```

Reclassifying density raster (converting to 0-1 scale)
```{r}
reclass_density <- matrix(c(
  0, 0.001789466, 0,
  0.001789466, 0.003259117, 0.25,
  0.003259117, 0.004418675, 0.5,
  0.004418675, 0.00615361, 0.75,
  0.00615361, Inf, 1
), ncol = 3, byrow = TRUE)

xjmax= 2.759048e-02 
xjmin  = 8.820881e-07 

dens_range <- (tracter - xjmin) / (xjmax - xjmin)
plot(dens_range) #this is the final layer that will be summed
```

Now that we have all of our layers in the correct format, we will sum them 
to conduct the suitability analysis

```{r}
summed <- (liquefaction_binned * 0.4) + (dens_range * 0.2) + (dists_range * 0.4)
summed
plot(summed, main = "Earthquake Risk Map")
plet(summed)
```
Now we need to reclassify our map so that it reads more clear, since right now 
the continuous distance raster and the rasterized liquefaction vector are slightly
unclear together.

```{r}
reclass_summed <- matrix(c(
  0, 0.2, 5,
  0.2, 0.4, 4,
  0.4, 0.6, 3,
  0.6, 0.8, 2,
  0.8, 1, 1 
), ncol = 3, byrow = TRUE)

hazard_risk <- classify(summed, reclass_summed)
plot(hazard_risk, col = colorRampPalette(c("red", "orange", "yellow"))(100))
plet(hazard_risk, col = colorRampPalette(c("red", "orange", "yellow"))(100))
```


Now that we have a classified map, we will figure out how many people live in 
high risk zones.

```{r}

tract_mean <- terra::extract(hazard_risk, merged, fun = mean, na.rm = TRUE)

merged$risk <- tract_mean[,2]

plot(merged, "risk", col = colorRampPalette(c("red", "orange", "yellow"))(100),
      main = "High Risk Tracts")

high_risk <- merged[merged$risk < 3, ]
plot(high_risk, add = TRUE, col = "purple")

sum(high_risk$Population)

```

Now we will conduct a sensitivity analysis by testing other weight combinations

To do this, we will create a function for reproducibility. 
```{r}
sensitivity <- function(weight1, weight2, weight3){
  summed <- liquefaction_binned * weight1 + dens_range * weight2 + dists_range * weight3
  hazard_risk <-  classify(summed, reclass_summed)
  tract_mean <- terra::extract(hazard_risk, merged, fun = mean, na.rm = TRUE)
  merged$risk <- tract_mean[,2]
  plot(merged, "risk", col = colorRampPalette(c("red", "orange", "yellow"))(100),
       main = "Tracts at High Risk",
       sub = paste0("Weights: Liquefaction = ", weight1, "; Density = ", weight2, 
                   "; Fault Distance = ", weight3))
  high_risk <- merged[merged$risk < 3, ]
  plot(high_risk, add = TRUE, col = "purple")
  return(sum(high_risk$Population))
}
```

What happens if we drop each variable from the sum?
```{r}

sensitivity(0.5, 0, 0.5) #density = 0
sensitivity(0, 0.5, 0.5) #liq = 0
sensitivity(0.5, 0.5, 0) #dist = 0
```


Testing equal weights and more weight on liquefaction
```{r}
sensitivity(0.33, 0.33, 0.33) # equal weights
sensitivity(0.5, 0.25, 0.25) #more weight on liquefaction
```

