---
title: "Working_Project"
author: "Wafaa Lawai"
date: "2026-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(terra)
library(readr)
library(classInt)
```

```{r}

# read in faults shapefile
faults <- vect("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/faults/shapefiles/FAM_Qt_Faults.shp")

# read in Alameda county shapefile
alameda <- vect("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/County_Boundary_-2312687860649160244/County_Boundary.shp")
```

Harmonize CRS & crop to county of interest
```{r}
faults <- project(faults, 'epsg:6417')
alameda <- project(alameda, 'epsg:6417')

alameda_faults <- crop(faults, alameda)
alameda_faults <- mask(faults, alameda)
```

Convert faults vector into a distance raster
```{r}
# createa  new raster to serve as distance raster
r <- rast(alameda, nrow = 180, ncol = 360)

dists <- distance(r, alameda_faults)

plot(dists)
plot(alameda_faults, add = TRUE, col = 'white')
```
Crop distance raster to county boundaries
```{r}
dists <- crop(dists, alameda)
dists <- mask(dists, alameda)

plot(dists)
plot(alameda_faults, add = TRUE, col = 'white')
```
Read in & tidy population data
```{r}
pop <- read.csv("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/alameda_pop.csv", header = FALSE)
head(pop)

pop <- t(pop)
pop <- as.data.frame(pop)

colnames(pop) <- c("Tract", "Population")
pop <- pop[-1,]

```

Create a separate ID column for population data
```{r}
pop$ID <- pop$Tract

for (i in 1:379) {
  pop$ID[i] <- parse_number(pop$ID[i])
  pop$Population[i] <- parse_number(pop$Population[i])
}

pop$Population <- as.numeric(pop$Population)

```

Area data

```{r}
area <- vect("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/AlamedaTracts.geojson")

for (i in 1:378) {
  area$DIST_NAME[i] <- parse_number(area$DIST_NAME[i])
}

```

Merge population data into spatvector with land area

```{r}
colnames(pop) <- c("Tract", "Population", "DIST_NAME")

tracts <- merge(area, pop, by = "DIST_NAME")
```

Calculate density

```{r}

tracts$density <- tracts$Population / tracts$area

# density <- density %>% 
#   mutate(density = Population / area)
```

boop
```{r}
h <- classIntervals(tracts$density, n = 5, style = 'quantile')

plot(tracts, "density", breakby = 'cases')
```

Turn this into a raster

```{r}

tracter <- rasterize(tracts, r, "density")

```


```{r}
liquefaction <- rast("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/liquefactionzones.tif")
plot(liquefaction)


reclass_liquefaction <- matrix(c(
  2, 0,
  1, 0.5,
  0, 1
), ncol = 2, byrow = TRUE)

liquefaction_binned <- classify(liquefaction, reclass_liquefaction)
plot(liquefaction_binned)


```
```{r}

reclass_density <- matrix(c(
  0, 0.001789466, 0,
  0.001789466, 0.003259117, 0.25,
  0.003259117, 0.004418675, 0.5,
  0.004418675, 0.00615361, 0.75,
  0.00615361, Inf, 1
), ncol = 3, byrow = TRUE)

xjmax= 2.759048e-02 
xjmin  = 8.820881e-07 

dens_range <- (tracter - xjmin) / (xjmax - xjmin)
plot(dens_range)


#density_binned <- classify(tracter, reclass_density)
#plot(density_binned)


#summed <- liquefaction_binned + density_binned  
#summed
#plot(summed)
```

```{r}
par(mfrow=c(2,2))
plot(density_binned)
plot(liquefaction_binned)
plot(density_binned)
plot(summed)
```


Distance raster binning/ reclass
```{r}
distmax= 14002.541
distmin  = 0.0047

dists_range <- (distmax - dists) / (distmax - distmin)
plot(dists_range)
```


```{r}
summed <- (liquefaction_binned * 0.4) + (dens_range * 0.2) + (dists_range * 0.4)
summed
plot(summed)
#plet(summed)



```
Reclassifying Risk

```{r}
reclass_summed <- matrix(c(
  0, 0.2, 5,
  0.2, 0.4, 4,
  0.4, 0.6, 3,
  0.6, 0.8, 2,
  0.8, 1, 1 
), ncol = 3, byrow = TRUE)

hazard_risk <- classify(summed, reclass_summed)
plet(hazard_risk, col = colorRampPalette(c("red", "orange", "yellow"))(100))
```

Overlaying population density tracts with risk and finding the approx population 
at high risk

```{r}
plot(hazard_risk)
plot(dens_range, add = TRUE)

tract_mean <- terra::extract(hazard_risk, tracts, fun = mean, na.rm = TRUE)

tracts$risk <- tract_mean[,2]

plot(tracts, "risk", col = colorRampPalette(c("red", "orange", "yellow"))(100))

high_risk <- tracts[tracts$risk >= 2 & tracts$risk <= 3.2, ]
plot(high_risk, add = TRUE, col = "purple")

sum(high_risk$Population)


```
Sensitivity Analysis (?)

```{r}

```




```{r}
roads <- vect("/Users/sonal/Documents/GitHub/esp106-project-Sonal-Wafaa/Street_Centerlines_-6175090382167386015/Street_Centerlines.shp")

roads <- project(roads, crs(summed))
```

```{r}

#plot(roads[roads$CLASS == 'Interstate'], add = TRUE)
#plot(roads[roads$CLASS == 'Freeway or Expressway'], add = TRUE)
#plot(roads[roads$CLASS == 'Minor Arterial'], add = TRUE)
#plot(roads[roads$CLASS == 'Principal Arterial'], add = TRUE)


roads_cropped <- crop(roads, alameda)
roads_masked <- mask(roads, alameda)



```





Combining roads dataset 

```{r}
plot(summed)

interstate <- roads_masked[roads_masked$CLASS == 'Interstate']
freeway <- roads_masked[roads_masked$CLASS == 'Freeway or Expressway']
arterial <- roads_masked[roads_masked$CLASS == 'Principal Arterial']
major_roads <- rbind(interstate, freeway, arterial)

r <- rast(extent = ext(dens_range), resolution = 100, crs = crs(dens_range)
)

plet(major_roads)
roads_raster <- rasterize(major_roads, r)

plet(roads_raster)

```


Sum with roads 

```{r}
res(dens_range) <- res(roads_raster)
res(density_binned) <- res(roads_raster)
res(liquefaction_binned) <- res(roads_raster)
summed_roads <- liquefaction_binned + density_binned + roads_raster + dens_range
plet(summed_roads)


```



